Ensemble Stability of Biological Networks

### 2014-10-15 tex/abstract-cited.tex
revise

%!TEX root = ../paper-abstract.tex
The analysis of dynamical systems that attempts to model chemical reaction, gene-regulatory, population, and ecosystem networks all rely on models having interacting components~\cite{RossCr2003,Alon2006,Palsson2006,HamidBolouri2008,Palsson2011a,Voit2012,Sauro2012}. When the details of these interactions are unknown for biological systems of interest, one effective approach is to study the dynamical properties of an ensemble of models determined by evolutionary constraints that may apply to all such systems~\cite{Gardner1970,May1972,Cohen1984,May1972a,Radius2014}. One such constraint is that of dynamical robustness, the probability of a stable system to remain stable under perturbation~\cite{WADDINGTON1942a,Wagner1997,Rutherford1998,VanNimwegen1999,Siegal2002,Bergman2003,Ciliberti2007b,Ciliberti2007,Draghi2010,Wagner2013}. Despite previous investigation, the relationship between dynamical robustness, an important functional characteristic of many biological systems, and network structure is poorly understood. Here we analyze the stability and robustness of a large class of dynamical systems and demonstrate that the most hierarchical network structures are the most robust. In particular, we determine the probability distribution of robustness over system connectivity and show that robustness is maximized by maximizing the number of links between strongly connected components of the graph representing the underlying system connectivity. We also demonstrate that permutation of strongly connected components is a fundamental symmetry of system robustness, which applies to networks of any number of components.  The classification of dynamical robustness based upon a purely topological property provides a fundamental organizing principle that can be used in the context of experimental validation to select among models that break or preserve network hierarchy. This result contributes to an explanation of the observation of hierarchical modularity across levels of the biological hierarchy~\cite{Zhao2006,Ravasz2002,Bhardwaj2010,Colm,Corominas-Murtra2013}.

>>>>

%!TEX root = ../paper-abstract.tex
The analysis of dynamical systems that attempts to model chemical reaction, gene-regulatory, population, and ecosystem networks all rely on models having interacting components~\cite{RossCr2003,Alon2006,Palsson2006,HamidBolouri2008,Palsson2011a,Voit2012,Sauro2012}. When the details of these interactions are unknown for biological systems of interest, one effective approach is to study the dynamical properties of an ensemble of models determined by evolutionary constraints that may apply to all such systems~\cite{Gardner1970,May1972,Cohen1984,May1972a,Radius2014}. One such constraint is that of dynamical robustness, the probability of a stable system to remain stable under perturbation~\cite{WADDINGTON1942a,Wagner1997,Rutherford1998,VanNimwegen1999,Siegal2002,Bergman2003,Ciliberti2007b,Ciliberti2007,Draghi2010,Wagner2013}. Despite previous investigation, the relationship between dynamical robustness, an important functional characteristic of many biological systems, and network structure is poorly understood. Here we analyze the stability and robustness of a large class of dynamical systems and demonstrate that the most hierarchical network structures are the most robust. In particular, we determine the probability distribution of robustness over system connectivity and show that robustness is maximized by maximizing the number of links between strongly connected components of the graph representing the underlying system connectivity. We also demonstrate that permutation of strongly connected components is a fundamental symmetry of dynamical robustness, which applies to networks of any number of components.  The classification of dynamical robustness based upon a purely topological property provides a fundamental organizing principle that can be used in the context of experimental validation to select among models that break or preserve network hierarchy. This result contributes to an explanation of the observation of hierarchical modularity across levels of the biological hierarchy~\cite{Zhao2006,Ravasz2002,Bhardwaj2010,Colm,Corominas-Murtra2013}.

### 2014-10-15 tex/discussion.tex
revise

%!TEX root = ../paper.tex
The relationship between structure and function is fundamental to biology. Structure-function relationships exist at every level of the biological hierarchy. In all such networks, the ability of systems to persist over long periods of time, is dependent upon their dynamical stability. Here we have demonstrated a structure-function relationship wherein biological networks that are more hierarchical are more robust and thus more likely to persist in the evolutionary process. This relationship transcends levels of the biological hierarchy in the sense that it is formulated mathematically and at a level of abstraction that allows its conclusions to apply independently to each of these levels.

Our development suggests two predictions that could be used to test the theory described here. One is that, if this theory is correct, any ensemble of systems where robustness has been selected for over a sufficiently long period of time, should exhibit a bias toward more hierarchical network topologies. In the short term, this prediction may be checked by looking at both metabolic and transcription factor networks, which have already been shown to display hierarchical structure, but whose dynamics have not been sufficiently well characterized to ascertain their robustness \cite{Zhao2006,Bhardwaj2010,Colm}. The other prediction of the theory outlined here that may be able to be checked is the converse statement that an ensemble of systems demonstrating a positive or negative bias in their network topology toward hierarchy may have been under selection for, or respectively against, robustness. Of course, in this case, factors other than selection for robustness would not be ruled out by this analysis alone.  In the long term, each of these predictions may be evaluated using experimental evolution by tracking the evolution of gene regulatory network topology in the context of environmental perturbations \cite{Leroi1994}.

The fact that such a relationship exists has the potential to unify our understanding of biological networks. In order to pursue this, it will be necessary to deepen our understanding of the relationship between dynamical robustness and the strongly connected components of the underlying network topology. One approach is to attempt to compute robustness for systems of larger size. Using existing methods, this would require large-scale Monte Carlo simulations. However, it may be possible to prove a bound on the robustness of individual strongly connected components, which could provide more definite knowledge and help to circumvent such expensive simulations. Nevertheless, the conservation of robustness with respect to nontrivial symmetries including the interchange of strongly connected components and permutation within strongly connected components suggests the existence of an evolutionary neutral space. A deeper mathematical characterization of the full symmetry groupoid of dynamical robustness may help to characterize a fundamental evolutionary constraint \cite{Golubitsky2006}.

>>>>

%!TEX root = ../paper.tex
The relationship between structure and function is fundamental to networks at every level of the biological hierarchy. Equally fundamental is the ability of systems to persist over long periods of time, which is dependent upon their dynamical robustness. Here we have demonstrated a structure-function relationship wherein biological networks that are more hierarchical are more robust and thus more likely to persist in the evolutionary process. Moreover, this result is formulated at a level of abstraction that allows its conclusions to apply independently to each of these levels.

Our development suggests two predictions that could be used to test the theory described here. One is that, if this theory is correct, any ensemble of systems where robustness has been positively selected over a sufficiently long period of time, should exhibit a bias toward more hierarchical network topologies. In the short term, this prediction may be checked by looking at both metabolic and transcription factor networks, which have already been shown to display hierarchical structure, but whose dynamics have not been sufficiently well characterized to ascertain their robustness \cite{Zhao2006,Bhardwaj2010,Colm}. The other prediction of the theory outlined here that may be able to be checked is the converse statement that an ensemble of systems demonstrating a positive or negative bias in their network topology toward hierarchy may have been under selection for, or respectively against, robustness. Of course, in this case, factors other than selection for robustness would not be ruled out by this analysis alone.  In the long term, each of these predictions may be evaluated using experimental evolution by tracking the evolution of gene regulatory network topology in the context of environmental perturbations \cite{Leroi1994}.

The fact that such a relationship exists has the potential to unify our understanding of biological networks. In order to pursue this, it will be necessary to deepen our understanding of the relationship between dynamical robustness and the underlying network topology. One approach is to attempt to compute robustness for systems of larger size. Using existing methods, this would require large-scale Monte Carlo simulations. However, it may be possible to prove a bound on the robustness of individual strongly connected components, which could provide more definite knowledge and help to circumvent such expensive simulations. Nevertheless, the conservation of robustness with respect to nontrivial symmetries including the interchange of strongly connected components and permutation within strongly connected components suggests the existence of an evolutionary neutral space. A deeper mathematical characterization of the full symmetry groupoid of dynamical robustness may help to characterize a fundamental evolutionary constraint \cite{Golubitsky2006}.

### 2014-10-15 tex/introduction.tex
draft results
revise

%!TEX root = ../paper.tex
The traditional approach taken in the study of chemical reaction, gene-regulatory, population, and ecosystem networks is to consider a particular example, derive a system of differential equations to model that example, try to fit the model to data and adjust the modeling assumptions along with parameter values until a good fit is achieved \cite{Meyer2014}. All of these models utilize essentially equivalent mathematical structure \ref{fig:biomodelexamples} \cite{RossCr2003,Alon2006,Palsson2006,HamidBolouri2008,Palsson2011a,Voit2012,Sauro2012}. Developing unified mathematical descriptions of all of these is one of the paramount goals of systems biology.

Recent work has demonstrated that as a result of \emph{sloppiness} in the dependence of qualitative dynamic phenomena on the geometry of parameter space that this approach allows for a large variety of models to fit the data \cite{Brown2003,Gutenkunst2007,Daniels2008a,Machta2013,Hines2014,Prabakaran2014,Tonsing2014}. In the face of uncertainty about the structure of such biological networks, to model the components under consideration as randomly interlinked becomes a reasonable approximation. This approach enables one to gain insight into what dynamical phenomena are possible to observe within a given class of dynamical systems, which is necessary to understand in order to determine whether or not a particular dynamical phenomenon should be regarded as unique or generic in the development and investigation of models applied to particular systems \cite{Gunawardena2013,Gunawardena2014}.

Indeed, investigating generic properties of a large class of dynamical systems was the approach taken by May in models of ecosystem dynamics \cite{Gardner1970,May1972}. The class of dynamical systems studied by May is not restricted to ecosystem dynamics and encompasses, among others, the dynamics of all of the biological networks represented in \ref{fig:biomodelexamples}. In particular, May conjectured on the basis of results from random matrix theory what eventually came to be referred to as the May-Wigner stability theorem \cite{Cohen1984,May1972a,Radius2014}. This relationship implies a relationship between a topological property, system connectivity, and a dynamical property, stability.

>>>>

%!TEX root = ../paper.tex
The traditional approach taken in the study of chemical reaction, gene-regulatory, population, and ecosystem networks is to derive a system of differential equations to model a particular biological network, attempt to fit that model to data and adjust the modeling assumptions along with parameter values until a good fit is achieved \cite{Meyer2014}. All of these models utilize essentially equivalent mathematical structures \ref{fig:biomodelexamples} \cite{RossCr2003,Alon2006,Palsson2006,HamidBolouri2008,Palsson2011a,Voit2012,Sauro2012}. Developing unified mathematical descriptions of all of these is one of the paramount goals of systems biology.

Recent work has demonstrated that as a result of inhomogeneity in the geometry of the sensitivity of the dynamics over the parameter space for such models, this approach allows for a large variety of models to fit the data \cite{Brown2003,Gutenkunst2007,Daniels2008a,Machta2013,Hines2014,Prabakaran2014,Tonsing2014}. In addition, there is often uncertainty about the very structure of such biological networks. In this context, it is crucial to gain insight into what dynamical phenomena are possible to observe within a given class of dynamical systems, which is necessary to understand in order to determine whether or not a particular dynamical phenomenon should be regarded as unique or generic in the development and investigation of models applied to particular systems \cite{Gunawardena2013,Gunawardena2014}. This can be achieved using a method common in statistical physics involving the consideration of an ensemble of systems that, in comparison to one another, appear to have components that are randomly interlinked.

Indeed, investigating generic properties of a large class of dynamical systems was the approach taken by May in models of ecosystem dynamics \cite{Gardner1970,May1972}. The class of dynamical systems studied by May is, however, not restricted to ecosystem dynamics and encompasses, among others, the dynamics of all of the biological networks represented in \ref{fig:biomodelexamples}. May conjectured on the basis of results from random matrix theory what eventually came to be referred to as the May-Wigner stability theorem \cite{Cohen1984,May1972a,Radius2014}. This implies a relationship between a topological property, system connectivity, and a dynamical property, stability.

### 2014-10-15 tex/supplement.tex
revise

This observation suggested that graph edit distance from \reffigscc (top panel), hierarchy, might provide a better characterization of system robustness. \reffigrobusthierarchy shows system robustness as a function of hierarchy, where the most hierarchical system has a hierarchy of 0. There is a monotonic correlation between the upper bound of robustness and hierarchy. \reffigconnectdist3D3x3 shows system robustness as a function of both hierarchy and connectivity. The monotonic correlation between hierarchy and robustness is refined by an underlying correlation between robustness and connectivity analogous to that of \reffigconnectcycle3D3x3.

>>>>

This observation suggested that graph edit distance from \reffigscc (top panel), hierarchy, might provide a better characterization of dynamical robustness. \reffigrobusthierarchy shows dynamical robustness as a function of hierarchy, where the most hierarchical system has a hierarchy of 0. There is a monotonic correlation between the upper bound of robustness and hierarchy. \reffigconnectdist3D3x3 shows dynamical robustness as a function of both hierarchy and connectivity. The monotonic correlation between hierarchy and robustness is refined by an underlying correlation between robustness and connectivity analogous to that of \reffigconnectcycle3D3x3.
